Rationality, Evolution \& the Environment

### 2015-08-21 paper/paper.tex
paper comments added

\textbf{some notes and remarks and todo's}

\begin{itemize}
\item micha: let's have a term for pairs (regret type, epistemic type); I'm inclined to use ``player
  type'', but that's what you reserved for regret type; maybe use \myalert{meta-type}?
\item 
\end{itemize}

@@

\section{Introduction / Motivation}
  part, of the evolutionary benefit of learning rules

\section{Rationality \& subjective preference}

The standard use of the word \textit{rationality} in economics and decision theory is to denote
the maximization of (subjective) expected utility. Since the work by vNM [cite], this stance
has been justified from the intuitive ''rationality'' of the axioms on the subjective
preference relation leading to the representability of the decision maker as an expected
utility maximizer.
While in the axiomatization by [vNM] the probability of the outcomes is objectively given a priori, Savage54 provides a larger set of axioms that represent the decision maker (DM) with both a subjective utility over outcomes \textit{and} a \textit{subjective} probability over states of the world. Savage's axioms constitute the foundation of subjective expected utility (SEU) theory, that is normally assumed to define the rational behavior of economic agents.\\

\begin{definition}[SEU-maximization] Let $S$ be a set of states of the world, $X$ a set of
  outcomes and $A$ a set of actions. Given a probability measure $P$ on $S$ and a subjective
  utility function $u:X \rightarrow \mathbb{R} $, an action $a^*: S \rightarrow X $ is
  SEU-maximizing if $a^* \in \text{argmax}_{a_i \in A} \int_S u(a^*(s))dP(s) $.
\end{definition}

@@

@@

\section{The model}

We consider an evolutionary model that is different, for the best of our knowledge, from all the standard models of evolutionary game theory. Given a game $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, a subjective utility over $G$ is a function $\theta: A \rightarrow \mathbb{R} $, where $ A:=\prod_{i\in N}A_{i} $ as usual. We denote by $ \Theta:= \mathbb{R}^{|A|} $ the set of all the subjective utilities over $G$. The model we introduce here is a multiple game model, that we also call \textit{multi-game} or \textit{meta-game}, where a \textit{class} $\mathcal{G}$ of fitness games is played in a population. All the models used in evolutionary game theory that we are aware of always consider a single game at a time. Our extension is meant to capture both a more general notion of environment and a more general notion of phenotype.\\ 
There are good reasons to claim that focusing on more general psychological mechanisms instead of on behavior itself as the phenotype under evolutionary selection can be more faithful and insightful for natural biology and ecology [Fawcett\&al.].
In all classical models the agents are represented as playing a single fixed game, and as genetically predetermined to play fixed strategies, that consequently are the only target of natural selection in those frameworks. Standard evolutionary game theory cannot take charge of this shift of perspective until it will keep modelling a single game at a time, and until it will keep focusing on observed behavior only. ''Instead we should expect animals to have evolved a set of psychological mechanisms which enable them to perform well on average across a range of different circumstances''[Fawcett\&al.]. [maybe add evolution of preferences here]

@@

@@

\subsection{Epistemic types}

@@

Given a player type and an epistemic type, there are different ways to produce a behavior. A huge body of experimental literature strongly suggests that agents are generally uncertainty (or ambiguity) averse [REF]. This pattern of behavior in the presence of ucertainty has been famously axiomatized by Gilboa and Schmeidler in a paper from 1989. The idea is that being uncertainty averse amounts to behaving in accordance with maximinimizing expected utility over a set of prior probabilities. Whenever the set of priors is a singleton, maximinization of expected utility coincides with maximization of expected utility. In line with empirical data, we assume that players in the population are uncertainty averse and conform to maxmin expected utility. The resulting behavior is then produced by maximinimizing over the subjective utility given by the player type and the set of priors given by the epistemic type. \\

@@

@@

\noindent We still treat them as distinct, because they will perform differently with respect to component-wise mutation (see below). Moreover, Fact \ref{fact:equivalence2x2} does not hold for $n \times n$ symmetric games in general.

@@

\begin{table}[]
\centering
\footnotesize
\begin{tabular}{ccccccccc}
  \hline
 & $\tuple{\tau^{reg}, \Delta(A)}$ 
 & $\tuple{\tau^{\pi}, \Delta(A)}$ 
 & $\tuple{\tau^{com}, \Delta(A)}$
 & $\tuple{\tau^{alt}, \Delta(A)}$
 & $\tuple{\tau^{reg}, \bar{\mu}}$ 
 & $\tuple{\tau^{\pi}, \bar{\mu}}$ 
 & $\tuple{\tau^{com}, \bar{\mu}}$
 & $\tuple{\tau^{alt}, \bar{\mu}}$ \\ 
  \hline
  $\tuple{\tau^{reg}, \Delta(A)}$ & 6.567 & 6.570 & 5.650 & 6.992 & 6.564 & 6.564 & 5.593 & 7.409 \\
  $\tuple{\tau^{\pi}, \Delta(A)}$ & 6.476 & 6.483 & 5.896 & 6.818 & 6.484 & 6.484 & 5.850 & 7.124 \\
  $\tuple{\tau^{com}, \Delta(A)}$ & 6.468 & 6.647 & 5.512 & 7.169 & 6.578 & 6.578 & 5.577 & 7.354 \\
  $\tuple{\tau^{alt}, \Delta(A)}$ & 5.968 & 5.923 & 5.363 & 6.685 & 5.975 & 5.975 & 5.086 & 6.973 \\
  $\tuple{\tau^{reg}, \bar{\mu}}$ & 6.908 & 6.918 & 5.988 & 7.456 & 6.929 & 6.929 & 5.934 & 7.783 \\
  $\tuple{\tau^{\pi}, \bar{\mu}}$ & 6.908 & 6.918 & 5.988 & 7.456 & 6.929 & 6.929 & 5.934 & 7.783 \\
  $\tuple{\tau^{com}, \bar{\mu}}$ & 6.529 & 6.680 & 5.445 & 7.276 & 6.542 & 6.542 & 5.521 & 7.440 \\
  $\tuple{\tau^{alt}, \bar{\mu}}$ & 6.450 & 6.337 & 5.772 & 6.978 & 6.457 & 6.457 & 5.479 & 7.500 \\
   \hline                          
\end{tabular}                      
\caption{Average payoff in simulations of 100,000
  randomly generated $n \times n$ symmetric games with $n$ randomly drawn from $\set{2, \dots, 10}$.}
\label{tab:ExpectedFitness_10x10}        
\end{table}

@@

@@

>>>>

\textbf{some general notes and remarks and todo's}

\begin{itemize}
\item micha: let's have a term for pairs (regret type, epistemic type); I'm inclined to use ``player
  type'', but that's what you reserved for regret type; maybe use \myalert{meta-type}?
\item can we reformulate the main result (proposition 1) in terms of ``strict dominance''
  and/or strict Nash equilibrium? sometimes the proof is not ideally precise because it
  actually shows something stronger, namely strict dominance, but everything is formulated in
  terms of stability; using dominance is also better for arguments about evolutionary dynamics:
  dynamics can be attracted by states that are not evolutionarily stable (even if there is an
  ESS), but dominated strategies will die out in any case;
\item should we give definitions for ESS and NSS in the main text?
\item should we include definitions of the replicator dynamic and/or the replicator mutator
  dynamic?
\item the most important issue with this text is still that we need to motivate why we care
  about non-probabilistic beliefs and/or security strategies; everything else will fall into
  place when that is done
\end{itemize}

@@

\section{Introduction / Motivation}
\label{sec:intr--motiv}

\todo[inline]{this is only a dummy introduction; rewrite eventually with a better focus on the
  results and main conceptual points we would like to drive home}

Evolutionary game theory has become an established philosophical tool for probing into
conceptual issues whose complexity requires mathematical modeling. However, most applications
of EGT look at evolutionary selection of behavior in a single \emph{stage game}. There are
actually two crude but common simplifications in this approach. Firstly, the stage game is
usually considered fixed, closed and immutable \citep[there are exceptions, of course, such
as][]{WordenLevin2007:Evolutionary-es, McKenzie-AlexanderSkymrs2012:Inventing-New-S}. Secondly,
what evolves are behavioral strategies for the stage game \citep[what][call the ``behavioral
gamit'']{FawcettHamblin2013:Exposing-the-be}.

It has consequently been argued that, instead of focusing on a fixed, immutable stage game and
the behavioral selection for that stage game alone, we should rather study the evolutionary
selection of \emph{choice mechanisms} that are adapted in a rich and variable environment
\citep[e.g.][]{FawcettHamblin2013:Exposing-the-be,McNamara2013:Towards-a-Riche}. However,
formal accounts of the adaptive value of choice mechanisms have focused, for the most part, on
the evolutionary benefit of learning rules
\citep[e.g.][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}. \todo{More
  references Harley 1981!!!, O Conor \dots}  Moreover, only little attention has been paid to
modeling the direct (game-theoretic) meta-competition between alternative choice mechanisms
\citep[see][for related criticism]{FawcettHamblin2013:Exposing-the-be}.

Here, we would therefore make two contributions to the literature on evolution of choice
mechanisms. On the conceptual side, we argue that paying attention to the evolution of
(subjective representations of) preferences matters. Since almost any reasonable learning or
decision-making process will make use of some (subjective, agent-internal (possibly
unconscious)) representation of the quality of choice options, the question of which such
representations are ecologically valuable (i.e., lead to high fitness) is central. Na\"ive
intuition might convince us that obviously preferences that veridically represent the object
fitness that evolutionary selection is based upon must surely be the best. We demonstrate that,
if we take into account how agents represent uncertainty, this na\"ive conclusion is premature.

On the methodological side, we introduce a conservative extension of standard EGT that allows
us to reuse notions such as evolutionary stability / evolutionary dynamics and yet helps us
study the evolutionary competition between choice mechanisms in a statistically variable
environment. Concretely, we look at the evolutionary competition of choice mechanisms in a
``meta-game'' that consists of the average expected payoffs of choice mechanisms when playing
arbitrary games (from a given class). In this sense, a meta-game captures (the modeller's)
assumptions about the relevant statistics of the environment in which evolutionary forces
operate.

Section~\ref{sec:rati--subj} introduces necessary concepts from formal decision and game
theory. Section \dots

\todo[inline]{add overview}

\section{Rationality, subjectivity  \& subjective regret}
\label{sec:rati--subj}

\todo[inline]{would it make sense to tighten and restructure this section a little bit? what
  about this: rationality says \emph{if} utils, beliefs and rational, then choose this; utils
  and beliefs are unobservable; in some case utils and preferences \emph{can} be tight to the
  environment (probability of a fair coin landing heads), in others no (probability of life on
  Mars); if tight to environment its frequencies and fitness that would pin down a norm for
  rational behavior, otherwise there's the danger of redescriptionism (just find some belief
  and preference that rationalizes observed behavior); \dots }

The standard definition of \textit{rationality} in economics and decision theory says that
choice is rational only if it maximizes (subjective) expected utility. Since the work by vNM
[cite], this view has been justified from the ``intuitive rationality'' of the axioms on the
subjective preference relation leading to the representability of the decision maker as an
expected utility maximizer. \todo{include axioms, or at least those that are mentioned as
  relevant later} While in the axiomatization by [vNM] the probability of the outcomes is
objectively given a priori, Savage54 provides a larger set of axioms that represent the
decision maker (DM) with both a subjective utility over outcomes \textit{and} a
\textit{subjective} probability over states of the world. Savage's axioms constitute the
foundation of subjective expected utility (SEU) theory, that is normally assumed to define the
rational behavior of economic agents.

\begin{definition}[SEU-maximization] Let $S$ be a set of states of the world, $X$ a set of
  outcomes and $A$ a set of actions. Given a probability measure $P$ on $S$ and a subjective
  utility function $u:X \rightarrow \mathbb{R} $, an action $a^*: S \rightarrow X $ is
  SEU-maximizing if $a^* \in \text{argmax}_{a_i \in A} \int_S u(a^*(s))dP(s) $. \todo{do we
    actually need outcomes for anything in this paper?}
\end{definition}

@@

\todo[inline]{personally, not a big fan of extended quotes; talks: yes, papers: don't know; is
  this dear to your heart? who's that Maclean anyway? ;-)}

@@

\section{The model}
\label{sec:model}

\todo[inline]{first two paragraphs here are partly redundant}

We consider an evolutionary model that is different, for the best of our knowledge, from all
the standard models of evolutionary game theory. Given a game
$ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, a subjective utility over $G$ is a function
$\theta: A \rightarrow \mathbb{R} $, where $ A:=\prod_{i\in N}A_{i} $ as usual. We denote by
$ \Theta:= \mathbb{R}^{|A|} $ the set of all the subjective utilities over $G$. The model we
introduce here is a multiple game model, that we also call \textit{multi-game} or
\textit{meta-game}, where a \textit{class} $\mathcal{G}$ of fitness games is played in a
population. All the models used in evolutionary game theory that we are aware of always
consider a single game at a time. \todo{better formulate less strongly} Our extension is meant
to capture both a more general notion of environment and a more general notion of phenotype.

There are good reasons to claim that focusing on more general psychological mechanisms instead
of on behavior itself as the phenotype under evolutionary selection can be more faithful and
insightful for natural biology and ecology [Fawcett\&al.].  In all classical models the agents
are represented as playing a single fixed game, and as genetically predetermined to play fixed
strategies, that consequently are the only target of natural selection in those
frameworks. Standard evolutionary game theory cannot take charge of this shift of perspective
until it will keep modelling a single game at a time, and until it will keep focusing on
observed behavior only. ''Instead we should expect animals to have evolved a set of
psychological mechanisms which enable them to perform well on average across a range of
different circumstances''[Fawcett\&al.]. [maybe add evolution of preferences here]

@@

\todo[inline]{can some of these definitions not be written as one? e.g., regret preference and
  regret type as one definition? seems more elegant to me, because it's less repetitive (``for
  a game ...'')}

@@

\subsection{Epistemic types}
\label{sec:epistemic-types}

\todo[inline]{I think we really need a better (and earlier) justification for these epistemic
  types; we could appeal to information available to agents; naturalness of situations in which
  we cannot attach numbers to our beliefs (life on Mars again) \dots}

@@

\todo[inline]{this paragraph seems to presuppose familiarity with maxi-minimization (can we
  write this with a hyphen, any chance?); we should rewrite this or introduce the notion immediately;}

Given a player type and an epistemic type, there are different ways to produce a behavior. A
huge body of experimental literature strongly suggests that agents are generally uncertainty
(or ambiguity) averse [REF]. This pattern of behavior in the presence of ucertainty has been
famously axiomatized by Gilboa and Schmeidler in a paper from 1989. The idea is that being
uncertainty averse amounts to behaving in accordance with maximinimizing expected utility over
a set of prior probabilities. Whenever the set of priors is a singleton, maximinization of
expected utility coincides with maximization of expected utility. In line with empirical data,
we assume that players in the population are uncertainty averse and conform to maxmin expected
utility. The resulting behavior is then produced by maximinimizing over the subjective utility
given by the player type and the set of priors given by the epistemic type.

  \todo[inline]{I would like to suggest to move Figure 2 next to Figure 1; it is already useful
  for example 1 and could be referred to there already; I would suggest to use subfigures to
  put both next to each other}

@@

\todo[inline]{should we explain here, or very soon, what the resulting types really are: regret
  minimizers, maximin players \dots}

@@

\noindent Two remarks. Firstly, Fact \ref{fact:equivalence2x2} does not hold for $n \times n$
symmetric games in general. Secondly, we treat behaviorally equivalent types as distinct,
because we assume that what is selected by biological evolution is not choice behavior as such,
but the mechanism which generates behavior. This matters in particular if we assume that
mutations affect different components of the behavior generating mechanisms (see below).

@@

\begin{table}[]
\centering
\footnotesize
\begin{tabular}{ccccccccc}
  \toprule
 & $\tuple{\tau^{reg}, \Delta(A)}$ 
 & $\tuple{\tau^{\pi}, \Delta(A)}$ 
 & $\tuple{\tau^{com}, \Delta(A)}$
 & $\tuple{\tau^{alt}, \Delta(A)}$
 & $\tuple{\tau^{reg}, \bar{\mu}}$ 
 & $\tuple{\tau^{\pi}, \bar{\mu}}$ 
 & $\tuple{\tau^{com}, \bar{\mu}}$
 & $\tuple{\tau^{alt}, \bar{\mu}}$ \\ 
  \midrule
  $\tuple{\tau^{reg}, \Delta(A)}$ & 6.567 & 6.570 & 5.650 & 6.992 & 6.564 & 6.564 & 5.593 & 7.409 \\
  $\tuple{\tau^{\pi}, \Delta(A)}$ & 6.476 & 6.483 & 5.896 & 6.818 & 6.484 & 6.484 & 5.850 & 7.124 \\
  $\tuple{\tau^{com}, \Delta(A)}$ & 6.468 & 6.647 & 5.512 & 7.169 & 6.578 & 6.578 & 5.577 & 7.354 \\
  $\tuple{\tau^{alt}, \Delta(A)}$ & 5.968 & 5.923 & 5.363 & 6.685 & 5.975 & 5.975 & 5.086 & 6.973 \\
  $\tuple{\tau^{reg}, \bar{\mu}}$ & 6.908 & 6.918 & 5.988 & 7.456 & 6.929 & 6.929 & 5.934 & 7.783 \\
  $\tuple{\tau^{\pi}, \bar{\mu}}$ & 6.908 & 6.918 & 5.988 & 7.456 & 6.929 & 6.929 & 5.934 & 7.783 \\
  $\tuple{\tau^{com}, \bar{\mu}}$ & 6.529 & 6.680 & 5.445 & 7.276 & 6.542 & 6.542 & 5.521 & 7.440 \\
  $\tuple{\tau^{alt}, \bar{\mu}}$ & 6.450 & 6.337 & 5.772 & 6.978 & 6.457 & 6.457 & 5.479 & 7.500 \\
   \bottomrule                         
\end{tabular}                      
\caption{Average payoff in simulations of 100,000
  randomly generated $n \times n$ symmetric games with $n$ randomly drawn from $\set{2, \dots, 10}$.}
\label{tab:ExpectedFitness_10x10}        
\end{table}

@@

\todo[inline]{text in this section and the next is rough and unpolished, only intended to have
  the basic results recorded, so that we can reshape the article around the basic results}

To see how preference types behave in evolutionary competition based on solitary decision
problems, we approximated, much in the spirit of meta-games, average accumulated payoffs
obtained in randomly generated solitary decision problems. For our purposes, a decision problem
$\tuple{\States, \Acts, \Utils}$ consists of a set of world states $\States$, a set of acts
$\Acts$, and a utility function $\Utils \mycolon \States \times \Acts \rightarrow
\mathbb{R}$.
We generated arbitrary decision problems by selecting numbers of states and acts
$n_\state, n_\act \in \set{2, \dots, 10}$ and then filling the utility table, so to speak, by
independently picking a number $\Utils(\state, \act) \in \set{0, 10}$ uniformly at random, just
as we did before with symmetric games. Unlike with two-player games, we need to also sample the
actual state of the world, which we selected uniformly at random from the available states in
the current decision problem. Player types, consisting of epistemic types $\bar{\mu}$ and
$\Delta(\States)$ and preference types $\tau^{reg}$ and $\tau^{\pi}$, give rise to behavior in
the same way as before. (It is obviously not possible to carry altruistic or competitive
preference types over to solitary decision making.) As before, we recorded the behavioral
decisions for each of the four relevant player types in each sampled decision problem, and
determined the actual payoff obtained, given the actual state. Average payoff over 100,000 decision
problems are given in Table~\ref{tab:SolitaryDecisions}.

\begin{table}
  \centering
  \begin{tabular}{cccc}
    \toprule
   $\tuple{\tau^{reg}, \Delta(A)}$ 
 & $\tuple{\tau^{\pi}, \Delta(A)}$ 
 & $\tuple{\tau^{reg}, \bar{\mu}}$ 
 & $\tuple{\tau^{\pi}, \bar{\mu}}$ 
 \\ \midrule
    6.318 & 6.237 & 6.661 & 6.661 \\ \bottomrule
  \end{tabular}
  \caption{Average payoffs over 100,000 simulated solitary decision problems}
  \label{tab:SolitaryDecisions}
\end{table}

Facts~\ref{fact:maxEU-minReg} and \ref{fact:equivalence2x2} still apply:
$\tuple{\tau^{reg}, \bar{\mu}}$ and $\tuple{\tau^{\pi}, \bar{\mu}}$ are behaviorally equivalent
in general, and $\tuple{\tau^{reg}, \Delta(A)}$ is behaviorally equivalent to the former two in
decision problems with 2 states and 2 acts. \todo{show this?} We see this in part in the
results in Table~\ref{tab:SolitaryDecisions} in that the averages for $\tuple{\tau^{reg},
  \bar{\mu}}$ and $\tuple{\tau^{\pi}, \bar{\mu}}$ are identical. But since we included decision
problems with more acts and more states as well, the averages for regret minimizers
$\tuple{\tau^{reg}, \Delta(A)}$ are not identical. They are, in fact, lower, but not as low as
those of the minimax choosers.

What that means is that basically every relevant result about game situations is also borne out
when reasoning about solitary decisions. Evolutionary selection based on objective fitness will
not select against regret-based subjective preferences, as these are indistinguishable from
veridical preferences when paired with probabilistic beliefs. But when paired with
non-probabilistic belief types, regret-based preferences actually outperform veridical
preferences. Consequently, if there is a chance, however small, that agents must fall back onto
non-probabilistic beliefs, evolution will actually positively select for non-veridical
regret-based preferences.

\subsection{Arbitrary probabilistic beliefs}
\label{sec:arbitr-prob-beli}

So far, we have assumed that epistemic types $\bar{\mu}$ hold an unbiased, flat belief. It is
worthwhile considering what happens when this assumption is relaxed and we allow agents with
probabilistic beliefs to make use of the full spectrum of probabilistic beliefs. It should be
clear that quality of beliefs directly impacts prospects of successful decision-making: if a
decision maker knows the actual state, or can put a high level of credence on the actual state,
accumulated fitness can be expected to be high. We see this, unsurprisingly, in numerical
simulations. We looked at the average payoff accumulated in 100,000 decision problems sampled
as before, except that for belief types with probabilistic beliefs we sampled a random
probabilistic belief $p$ (from an unbiased Dirichlet distribution) and chose the actual world
state with probability weights $p$. This effectively implements a bias for beliefs of the
decision maker that tend to put more weight on the actual state (although the procedure does
admittedly perverse the natural chicken-and-egg logic in this case that the actual world state
should come first and beliefs be a function of it). As a result, the average payoffs of types
$\tuple{\tau^{reg}, \bar{\mu}}$ and $\tuple{\tau^{\pi}, \bar{\mu}}$ are still the exact same
(because we compare what agents with different preference types would do under the same
beliefs; we are not interested in statistical fluctuations given one type better beliefs by
pure happenstance), but with $7.125$ notably higher than before. In effect, better
probabilistic beliefs lead to better decisions. If agents can learn, reason and extrapolate to
form better probabilistic beliefs, that will help them whenever they take their beliefs into
account.

But this is orthogonal to the arguments in this paper, where the focus is on possibilities for
the evolution of non-veridical preferences. We know from Fact~\ref{fact:maxEU-minReg} that
regret types and veridical preference types, when paired with probabilistic belief types that
maximize/minimize expectations, are behaviorally equivalent, \emph{no matter what they belief},
as long as they belief the same. So, if learning, insight and statistical knowledge of a
recurrent situation \emph{can} be brought to bear, this will not make evolution select against
regret-based preferences. If, on the other hand, agents have no basis for probabilistic
beliefs, then we have shown that there are general and basic circumstances in which
regret-based preferences can actually be selected, despite their non-veridicality.

@@

  part, on the evolutionary benefit of learning rules
